Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2

### 2025-03-11 submitted/go3-author-sub.tex
submitted version Dec 6

\title{Explicit Scale Simulation for analysis of RNA-sequencing with
ALDEx2}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (size) of the biological system. Typically a statistical
normalization removes unwanted technical variation in the data or the
parameters of the model to enable differential abundance analyses. We
recently showed that all normalizations make implicit assumptions about
the unmeasured system scale and that errors in these assumptions can
dramatically increase false positive and false negative rates. We
demonstrated that these errors can be mitigated by accounting for
uncertainty using a \emph{scale model}, which we integrated into the
ALDEx2 R package. This article provides new insights focusing on the
application to transcriptomic analysis. We provide transcriptomic case
studies demonstrating how scale models, rather than traditional
normalizations, can reduce false positive and false negative rates in
practice while enhancing the transparency and reproducibility of
analyses. These scale models replace the need for dual cutoff approaches
often used to address the disconnect between practical and statistical
significance. We demonstrate the utility of scale models built based on
known housekeeping genes in complex metatranscriptomic datasets. Thus
this work provides guidance on how to incorporate scale into
transcriptomic data sets.
\end{abstract}

>>>>

\title{Explicit Scale Simulation for analysis of RNA-sequencing with
ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin D. Silverman$^{2,3,4,5}$}
\date{\today}

\begin{document}
\maketitle

$^1$Department of Biochemistry, University of Western Ontario
$^2$Department of Population Health Sciences, Geisinger, Danville, PA
$^3$Department of Medicine, Pennsylvania State University
$^4$Institute for Computational and Data Science, Pennsylvania State University
$^5$Department of Statistics, Pennsylvania State University

\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (size) of the biological system. Typically a statistical
normalization removes unwanted technical variation in the data or the
parameters of the model to enable differential abundance analyses. We
recently showed that all normalizations make implicit assumptions about
the unmeasured system scale and that errors in these assumptions can
dramatically increase false positive and false negative rates. We
demonstrated that these errors can be mitigated by accounting for
uncertainty using a \emph{scale model}, which we integrated into the
ALDEx2 R package. This article provides new insights focusing on the
application to transcriptomic analysis. We provide transcriptomic case
studies demonstrating how scale models, rather than traditional
normalizations, can reduce false positive and false negative rates in
practice while enhancing the transparency and reproducibility of
analyses. These scale models replace the need for dual cutoff approaches
often used to address the disconnect between practical and statistical
significance. We demonstrate the utility of scale models built based on
known housekeeping genes in complex metatranscriptomic datasets. Thus
this work provides guidance on how to incorporate scale into
transcriptomic data sets.
\end{abstract}
